Tabular Constraint Learning

### 2016-04-13 Paper/ECAI/tabular.tex
Fixes

\newcommand{\ecrank}[2]{\ensuremath{#1 = \mathit{RANK}(#2)}}
\newcommand{\ecfkey}[2]{\ensuremath{#1 \rightarrow #2}}
\newcommand{\ecalldiff}[1]{\ensuremath{\mathit{ALLDIFFERENT}(#1)}}

@@

\begin{abstract}
  abstract
\end{abstract}
\section{Introduction}
Imagine you have received a CSV file. Upon examining the content it becomes fairly clear it was made in Excel. Even the names of the fields such as Total or Avg indicated the usage of the standard functions. Having a closer look, you could deduce the functional dependencies in the table that look as in Figure \ref{fig:main_example}. Provided that a spreadsheet and the numbers in it are small, and dependencies are not complicated, one can examine the spreadsheet and find all connections. But wouldn't it be good to do it automatically?

@@

Of course, it is stated semi-informally here, but it already allows us to provide an intuition on the task. We know that often people work in Excel with ranges, e.g. if we look at the Table 1  and at the first row in the Table 2 in Figure \ref{fig:main_example}, it is clear that one could have selected the first columns, clicked on the \textit{SUM} function and then just used the drag fill handle to apply the formula to the other columns. Similarly, that would work for the second (\textit{Average}), third (\textit{Min})  and forth (\textit{Max}) rows in Table 2. I.e., all operations are vectorized and make use of the standard Excel functions.

@@

An \textit{Excel constraint} is a triple \textit{(\CName, \CSignature, \CFunction)}. Let us elaborate on each of them. \textit{\CName} is the textual name of the constraint together with its variable names. \textit{\CSignature} is a set of constraints specifying the properties of the group assignments, corresponding to the variables in \CName, such as their types, e.g., requiring them to be integers or constraining the sizes, e.g. the length of vectors in the arguments must be equal. These are constraints on the group meta-information not on the actual group content. \textit{\CFunction} is a set of constraints specifying that the data in the subgroups satisfies the function. For example, an excel constraint \textit{rank} has \CName \textit{Y = RANK(X)}, where $X$ and $Y$ are the variables; its \CSignature is: the group $G_X$ (associated with X) is numeric, $G_Y$ is integer and the length of vectors in $G_X$ is the same as in $G_Y$; its \CFunction is the following constraint: a pair of vectors $X,Y$ is a solution iff $X \in G_X, Y \in G_Y$ and each value in $X$ has the rank (possibly with ties) specified in $Y$.

@@

A \textit{tabular solution} $S$ to an Excel constraint $C$ is a tuple $(X_1, \dots, X_n)$ where $n$ is the number of variables in $C$ and each $X_i$ is a subgroup in \groups, such that $(X_1,\dots,X_n)$ satisfies both the \CSignature and \CFunction of the constraint $C$.

\section{Problem Statement}
In the previous section we introduced the problem of tabular constraint learning informally using the example in Figure \ref{fig:main_example}. Here we formalize the statement in terms of Excel constraints and group assignments as follows:

\begin{minipage}[c]{14em}
  \vspace{5pt}
  \begin{tabular}{ll}
    \multicolumn{2}{l}{{\textbf{Tabular Constraint Learning Problem}}}\\
    \vspace{-4pt}
    &\\
    \textbf{Given:}& the set of all groups $\groups$ and of Excel constraints $\constraints$\\
    \textbf{Find:}&  all tabular solution $S$ in \groups for each constraint $c$ \constraints \\ 
  \end{tabular}
  \vspace{6pt}
\end{minipage}

  The key observation here is that essentially the problem is a constraint enumeration problem, where each constraint is independent. This property comes from fact that in Excel each formula is applied based on the existing data in the cells. This allows learning constraints independently of each other by examining constraint satisfaction on the groups.

@@

\begin{minipage}[c]{14em}
  \vspace{5pt}
  \begin{tabular}{ll}
    \multicolumn{2}{l}{{\textbf{Condensed Tabular Constraint Learning Problem}}}\\
    \vspace{-4pt}
    &\\
    \textbf{Given:}& the set of all groups $\groups$ and of Excel constraints $\constraints$,\\ 
    & a constraint specificity DAG \dependencies \\
    \textbf{Find:}& all the most specific wrt \dependencies solutions $S$ in \groups\\ 
    & for each $c$ in \constraints \\
  \end{tabular}
  \vspace{6pt}
\end{minipage}

@@

\begin{algorithm}[thb]
  \begin{algorithmic}
    \footnotesize
    \State \textbf{Input:} $\groups$ -- groups, $\constraints$ -- Excel constraints, \dependencies -- dependency graph 
    \State \textbf{Output:} $S$ -- learned constraints with their satisfaction assignments
    \State $S \gets \emptyset$ \Comment{The set of solutions}
  \For{$c \in \constrainttorder(\constraints,\dependencies)$} 
  \State $v_1, ..., v_n =$ variables of $c$ 
    \For{$v_1{:}~G_1, \dots, v_n{:}~G_n \in \generategroups(c, G, S, \dependencies)$}
      \State $S \gets S \cup \findassignment(c, v_1{:}~G_1, \dots, v_n{:}~G_n, S, \dependencies)$
    \EndFor
  \EndFor\\
\Return $S$
\end{algorithmic}
\caption{Tabular constraint learning}
\label{algo:tcl}
\end{algorithm}
Essentially, our Algorithm \ref{algo:tcl} has three steps: constraint ordering, candidate group generation and subgroup satisfaction search, which is in line with the ``generate-and-test'' paradigm that is well-known in AI \cite{whaisasp}. Let us elaborate on each step in detail.

\samuel{Running example (RANK?)}
\subsection{Constraint ordering} $\constrainttorder(\constraints,\dependencies)$ uses the DAG \dependencies in Figure \ref{fig:learning_order} in the following way: all constraints are generated in the topological order of \dependencies.
Solid edges point from more general to more specific constraints.
Therefore, if there is a solid edge from $C_1$ to $C_2$, then a set of subgroups satisfying $C_2$ must contain the subgroups satisfying $C_1$.

@@

@@

\subsection{Constraints}
As emphasized before, the set of Excel constraints has a partial order in which they should be learned. For certain constraints this order does not matter, such as \textit{rank} or \textit{product}. For many others they have to be learned in the order, which is specified as a DAG in Figure \ref{fig:learning_order} (if a constraint is not in the graph, it is independent). This figure also specifies a potential parallelization of the computation, since each connected component is independent of the others.

@@

\begin{table*}
  \centering
  \begin{tabularx}{\textwidth}{XXX}
    \textbf{\CName} & \textbf{\CSignature} & \textbf{\CFunction}\\
    ALLDIFFERENT(X) & X is integer or textual & All elements in X are different \\
    %X = Y & & \\
    FK $\rightarrow$ PK & FK and PK have the same type; they originate from different tables and PK is all-different & all values in FK exist in PK \\
    % TODO check out
    R = LOOKUP-PRODUCT(FV, FK=PK, PV) & & \\
    FV = FUZZY-LOOKUP(FK, PK, PV) & & \\
    FV = LOOKUP(FK, PK, PV) & & \\
    PERMUTATION(X) & & \\
    R = O1 * O2 & & \\
    R = PROJECT(P) & & \\
    Y = RANK(X) & Y and X have the same length & Y is the rank of X (including ties)\\
    A = PREV(A) + P - N & A, P and N have the same length and their length is at least 2 & *\\
    SERIES(X) & & \\
    Y = SUM(X, col) & & Y is the result of summing each column in X \\
    Y = SUM(X, row) & & Y is the result of summing each row in X \\
    R = SUMIF(FK=PK, V) & & \\
    R = SUMPRODUCT(O1, O2) & & \\

@@

  \end{tabularx}
  \caption{Excel Tabular Constraints \sergey{Luc, we need your comments here on the notation of $X,Y$ and $G_X,G_Y$}}
  \label{table:constraints}
\end{table*}

@@

In this section we focus on \textit{functional} constraints that could be used in Excel, ignoring constraints such as all-different or foreign-key.

@@

\begin{figure}
  \label{fig:sol_example}
  {\small
    \begin{align*}
      & SERIES(\range{T_{1}}{\rangeall}{1}) \\
      & \range{T_{1}}{\rangeall}{8} = RANK(\range{T_{1}}{\rangeall}{7}) \\
      & \range{T_{2}}{1}{\rangeall} = SUM(\range{T_{1}}{\rangeall}{\rangeto{3}{7}}, col) \\
      & \range{T_{6}}{\rangeall}{2} = SUM(\range{T_{1}}{\rangeall}{\rangeto{3}{6}}, col) \\
      & \range{T_{1}}{\rangeall}{7} = SUM(\range{T_{1}}{\rangeall}{\rangeto{3}{6}}, row) \\
      & \range{T_{2}}{2}{\rangeall} = AVERAGE(\range{T_{1}}{\rangeall}{\rangeto{3}{7}}, col) \\
      & \range{T_{2}}{3}{\rangeall} = MAX(\range{T_{1}}{\rangeall}{\rangeto{3}{7}}, col) \\
      & \range{T_{2}}{4}{\rangeall} = MIN(\range{T_{1}}{\rangeall}{\rangeto{3}{7}}, col) \\
      & \range{T_{1}}{\rangeall}{10} = SUMIF(\range{T_{5}}{\rangeall}{1}=\range{T_{1}}{\rangeall}{2}, \range{T_{5}}{\rangeall}{2}) \\
      & \range{T_{1}}{\rangeall}{11} = MAXIF(\range{T_{5}}{\rangeall}{1}=\range{T_{1}}{\rangeall}{2}, \range{T_{5}}{\rangeall}{2}) \\
      & \range{T_{4}}{\rangeall}{3} = LOOKUP(\range{T_{4}}{\rangeall}{2}, \range{T_{1}}{\rangeall}{1}, \range{T_{1}}{\rangeall}{2}) \\
      & \range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}
  }
  \caption{Expected constraints case study}
\end{figure}

>>>>

\newcommand{\eccalc}[2]{\ensuremath{#1 = #2}}
\newcommand{\ecrank}[2]{\eccalc{#1}{\mathit{RANK}(#2)}}
\newcommand{\ecfkey}[2]{\ensuremath{#1 \rightarrow #2}}
\newcommand{\ecalldiff}[1]{\ensuremath{\mathit{ALLDIFFERENT}(#1)}}
\newcommand{\eclookupf}[4]{\ensuremath{\mathit{LOOKUP}_{#4}(#1, #2, #3)}}
\newcommand{\eclookup}[4]{\eccalc{#1}{\eclookupf{#2}{#3}{#4}{}}}
\newcommand{\eclookupprod}[5]{\eccalc{#1}{#2 \times \eclookupf{#3}{#4}{#5}{}}}
\newcommand{\eclookupfuzzy}[4]{\eccalc{#1}{\eclookupf{#2}{#3}{#4}{fuzzy}}}
\newcommand{\ecperm}[1]{\ensuremath{\mathit{PERMUTATION}(#1)}}
\newcommand{\ecseries}[1]{\ensuremath{\mathit{SERIES}(#1)}}
\newcommand{\ecprod}[3]{\eccalc{#1}{#2 \times #3}}
\newcommand{\ectotal}[3]{\eccalc{#1}{\mathit{PREV}(#1) + #2 - #3}}
\newcommand{\ecproj}[2]{\eccalc{#1}{\mathit{PROJECT}(#2)}}
\newcommand{\ecaggc}[3]{\eccalc{#2}{\mathit{#1}(#3, col)}}
\newcommand{\ecaggr}[3]{\eccalc{#2}{\mathit{#1}(#3, row)}}
\newcommand{\ecsumc}[2]{\eccalc{#1}{\mathit{SUM}(#2, col)}}
\newcommand{\ecsumr}[2]{\eccalc{#1}{\mathit{SUM}(#2, row)}}
\newcommand{\ecaggif}[5]{\eccalc{#2}{\mathit{#1IF}(#3, #4, #5)}}
\newcommand{\ecsumif}[4]{\eccalc{#1}{\mathit{SUMIF}(#2, #3, #4)}}
\newcommand{\ecsumprod}[3]{\eccalc{#1}{\mathit{SUMPRODUCT}(#2, #3)}}

@@

\begin{abstract}
  abstract
\end{abstract}

\tias{Use spreadsheet instead of Excel}
\tias{Spreadsheet is an instance of tabular data, but not the only one?}

\tias{Rewrite}
\section{Introduction}
Imagine you have received a CSV file. Upon examining the content it becomes fairly clear it was made in Excel. Even the names of the fields such as Total or Avg indicated the usage of the standard functions. Having a closer look, you could deduce the functional dependencies in the table that look as in Figure \ref{fig:main_example}. Provided that a spreadsheet and the numbers in it are small, and dependencies are not complicated, one can examine the spreadsheet and find all connections. But wouldn't it be good to do it automatically?

@@

Of course, it is stated semi-informally here, but it already allows us to provide an intuition on the task. We (know) that often people work in spreadsheets with ranges, e.g. if we look at the Table 1  and at the first row in the Table 2 in Figure \ref{fig:main_example}, it is clear that one could have selected the first columns, clicked on the \textit{SUM} function and then just used the drag fill handle to apply the formula to the other columns. Similarly, that would work for the second (\textit{Average}), third (\textit{Min})  and forth (\textit{Max}) rows in Table 2. I.e., all operations are vectorized and make use of the standard spreadsheet functions.

@@

\tias{Use generic constraint}
A \textit{Spreadsheet constraint} is a triple \textit{(\CName, \CSignature, \CFunction)}. Let us elaborate on each of them. \textit{\CName} is the textual name of the constraint together with its variable names. \textit{\CSignature} is a set of constraints specifying the properties of the group assignments, corresponding to the variables in \CName, such as their types, e.g., requiring them to be integers or constraining the sizes, e.g. the length of vectors in the arguments must be equal. These are constraints on the group meta-information not on the actual group content. \textit{\CFunction} is a set of constraints specifying that the data in the subgroups satisfies the function. For example, an excel constraint \textit{rank} has \CName \textit{Y = RANK(X)}, where $X$ and $Y$ are the variables; its \CSignature is: the group $G_X$ (associated with X) is numeric, $G_Y$ is integer and the length of vectors in $G_X$ is the same as in $G_Y$; its \CFunction is the following constraint: a pair of vectors $X,Y$ is a solution iff $X \in G_X, Y \in G_Y$ and each value in $X$ has the rank (possibly with ties) specified in $Y$.

@@

A \textit{tabular solution} $S$ to an spreadsheet constraint $C$ is a tuple $(X_1, \dots, X_n)$ where $n$ is the number of variables in $C$ and each $X_i$ is a subgroup in \groups, such that $(X_1,\dots,X_n)$ satisfies both the \CSignature and \CFunction of the constraint $C$.

\section{Problem Statement}
In the previous section we introduced the problem of tabular constraint learning informally using the example in Figure \ref{fig:main_example}. Here we formalize the statement in terms of spreadsheet constraints and group assignments as follows:

\begin{minipage}[c]{14em}
  \vspace{5pt}
  \begin{tabular}{ll}
    \multicolumn{2}{l}{{\textbf{Tabular Constraint Learning Problem}}}\\
    \vspace{-4pt}
    &\\
    \textbf{Given:}& the set of all groups $\groups$ and of spreadsheet constraints $\constraints$\\
    \textbf{Find:}&  all tabular solution $S$ in \groups for each constraint $c$ \constraints \\ 
  \end{tabular}
  \vspace{6pt}
\end{minipage}

  The key observation here is that essentially the problem is a constraint enumeration problem, where each constraint is independent. This property comes from fact that in spreadsheets each formula is applied based on the existing data in the cells. This allows learning constraints independently of each other by examining constraint satisfaction on the groups.

@@

\begin{minipage}[c]{14em}
  \vspace{5pt}
  \begin{tabular}{ll}
    \multicolumn{2}{l}{{\textbf{Condensed Tabular Constraint Learning Problem}}}\\
    \vspace{-4pt}
    &\\
    \textbf{Given:}& the set of all groups $\groups$ and of spreadsheet constraints $\constraints$,\\ 
    & a constraint specificity DAG \dependencies \\
    \textbf{Find:}& all the most specific wrt \dependencies solutions $S$ in \groups\\ 
    & for each $c$ in \constraints \\
  \end{tabular}
  \vspace{6pt}
\end{minipage}

@@

\begin{algorithm}[thb]
  \begin{algorithmic}
    \footnotesize
    \State \textbf{Input:} $\groups$ -- groups, $\constraints$ -- Spreadsheet constraints, \dependencies -- dependency graph 
    \State \textbf{Output:} $S$ -- learned constraints with their satisfaction assignments
    \State $S \gets \emptyset$ \Comment{The set of solutions}
  \For{$c \in \constrainttorder(\constraints,\dependencies)$} 
  \State $v_1, ..., v_n =$ variables of $c$ 
    \For{$v_1{:}~G_1, \dots, v_n{:}~G_n \in \generategroups(c, G, S, \dependencies)$}
      \State $S \gets S \cup \findassignment(c, v_1{:}~G_1, \dots, v_n{:}~G_n, S, \dependencies)$
    \EndFor
  \EndFor\\
\Return $S$
\end{algorithmic}
\caption{Tabular constraint learning}
\label{algo:tcl}
\end{algorithm}
Essentially, our Algorithm \ref{algo:tcl} has three steps: constraint ordering, candidate group generation and subgroup satisfaction search, which is in line with the ``generate-and-test'' paradigm that is well-known in AI \cite{whaisasp}. Let us elaborate on each step in detail.

\samuel{Running example (RANK?)}

\subsection{Constraint ordering} $\constrainttorder(\constraints,\dependencies)$ uses the DAG \dependencies in Figure \ref{fig:learning_order} in the following way: all constraints are generated in the topological order of \dependencies.
Solid edges point from more general to more specific constraints.
Therefore, if there is a solid edge from $C_1$ to $C_2$, then a set of subgroups satisfying $C_2$ must contain the subgroups satisfying $C_1$.

@@

\dependencies thus describes a partial order in which constraints should be learned.
Some constraints have no dependencies (ommitted from the figure) and can be learned independently.
Generally, a parallel version of the algorithm could learn each connected component independently, only taking into account the partial ordering within components.

@@

\subsection{Constraints}
% As emphasized before, the set of Excel constraints has a partial order in which they should be learned. For certain constraints this order does not matter, such as \textit{rank} or \textit{product}. For many others they have to be learned in the order, which is specified as a DAG in Figure \ref{fig:learning_order} (if a constraint is not in the graph, it is independent). This figure also specifies a potential parallelization of the computation, since each connected component is independent of the others.

@@

\begin{table*}
  \centering
  \begin{tabularx}{\textwidth}{XXX}
    \textbf{\CName} & \textbf{\CSignature} & \textbf{\CFunction}\\
    \ecalldiff{V_x} & X is integer or textual & All elements in X are different \\
    %X = Y & & \\
    \ecfkey{V_{fk}}{V_{pk}} & FK and PK have the same type; they originate from different tables and PK is all-different & all values in FK exist in PK \\
    % TODO check out
    \eclookupprod{V_r}{V_1}{V_{fk}}{V_{pk}}{V_{val}} & & \\
    \eclookup{V_r}{V_{fk}}{V_{pk}}{V_{val}} & & \\
    \eclookupfuzzy{V_r}{V_{fk}}{V_{pk}}{V_{val}} & & \\
    \ecperm{V_x} & & \\
    \ecprod{V_r}{V_1}{V_2} & & \\
    \ecproj{V_r}{G_x} & & \\
    \ecrank{V_r}{V_1} & Y and X have the same length & Y is the rank of X (including ties)\\
    \ectotal{V_r}{V_{pos}}{V_{neg}} & A, P and N have the same length and their length is at least 2 & *\\
    \ecseries{V_1} & & \\
    \ecsumc{V_r}{G_1} & & Y is the result of summing each column in X \\
    \ecsumr{V_r}{G_1} & & Y is the result of summing each row in X \\
    \ecsumif{V_r}{V_{fk}}{V_{pk}}{V_{val}} & & \\
    \ecsumprod{V_r}{V_1}{V_2} & & \\

@@

  \end{tabularx}
  \caption{Spreadsheet Constraints \sergey{Luc, we need your comments here on the notation of $X,Y$ and $G_X,G_Y$}}
  \label{table:constraints}
\end{table*}

@@

In this section we focus on \textit{functional} constraints that could be used in spreadsheets, ignoring constraints such as all-different or foreign-key.

@@

\begin{figure}
  \label{fig:sol_example}
  {\small
    \begin{align*}
      & SERIES(\range{T_{1}}{\rangeall}{1}) \\
%
      & \ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}} \\
%
      & \ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecsumc{\range{T_{6}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      & \ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      & \ecaggc{AVERAGE}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggc{MAX}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggc{MIN}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggif{SUM}{\range{T_{1}}{\rangeall}{10}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      & \ecaggif{MAX}{\range{T_{1}}{\rangeall}{11}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      & \eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}} \\
%
      & \range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}
  }
  \caption{Expected constraints case study}
\end{figure}
